Rapid, (semi-?)Supervised Generation of Novel Percussive Sounds With A Tractable Virtual Synthesizer

### 2020-01-02 NIME_PAPER/paper.tex
Updates from Overleaf

\usepackage{float}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{todonotes}
\usepackage{etoolbox}
\usepackage[inline]{enumitem}
\begin{document}

@@

\section{Introduction}
\subsection{Terminology}
I'd like to define terms such as VSTs, audio effects, Synthesizers, paramters and presets here. I throw these terms around a lot and I think some might be confused by their meaning.
\subsection{Background and Motivation}
The rise of Digital Audio Workstations (DAW) \cite{leider2004digital} and Virtual Studio Technology (VST) based plug-ins \cite{tanev2013virtual} have rapidly transformed the sonic and material landscape of music production in the recent years. Coupled with this rise in popularity is a vast array of commercial products and services dedicated to satiating the need of amateur and professional music producers for unique sounds; most commonly via audio samples: one-shot drum samples, long sustained notes (commonly referred to as pads or textures), and loops(percussive or melodic) are common deliverables. Two notable examples of these commercial services are \textit{loopmasters}\footnote{loopmasters.com} and \textit{splice.com}\footnote{splice.com}. Furthermore, VST plug-ins can emulate complex audio synthesizers and effects, which some producers may find daunting or time consuming to work with from scratch. In many cases, VST plug-in vendors or unaffiliated enthusiasts sell additional presets for these plugins, targeted towards producers who do not have the time or interest in creating their own. Producers may further modify these presets until their desired sound is reached.\\
Our work is motivated by the idea of finding new, convenient methods for the expansion of a music producer's library of sounds. Primarily with generation of novel, one-shot audio samples but also with automated search and creation of presets for virtual synthesizers that a producer might find  interesting. Using the generation of short, percussive audio samples as a starting point, this project is a proof of concept for promising avenues towards our motivational goal.\\

@@

\subsection{Related Work and Contradistinctions}
\label{related}
Numerous deep, neural network models have been proposed and utilized for the purpose of signal generation in recent years. WaveGans and WaveNet have been subject to significant improvements and experiments since their proposal \cite{nsynth2017} and recently, Variational AutoEncoders (VAE's) have been utilized for generation of short percussive samples \cite{aouameur2019neural,ramires2019timbfeat}. In this work however, we opt to use digital signal processing methods to create a virtual synthesizer for the generation of audio signals, as it provides several unique advantages:
\begin{enumerate}[label=\roman*]
  \item Fast, offline rendering of audio with no relience on GPU, currently not possible with state of the art models such as parallel WaveGan \cite{yamamoto2019parallel} and parallell WaveNet \cite{oord2017parallel}. 
  \item Rendering at high sampling rates with little performance issues: Performance speed being a common issue, the standard sampling rate in most audio generation work utilizing neural networks appears to be under 24 khz \cite{yamamoto2019parallel,oord2017parallel,aouameur2019neural,ramires2019timbfeat}. However, a significant number of untrained human ears can detect a change in quality of audio between sampling rates of 192 khz and the industry standard of 44.1 khz \cite{reiss2016meta}, with a dramatic increase in quality detection after training. Therefore we can safely assume that most producers would prefer their audio samples have sampling rates of 44.1 khz or higher. 
  \item Neural networks are often viewed as unexplainable black box solutions. Some models such as VAE's can learn an underlying latent space of parameters and capture the "essence" of the different labels in a dataset. However, these spaces are learned in an unsupervised manner and must be manually analysed, perhaps extensively, before they can be understood \cite{esling2018generative}. The use of a virtual synthesizer for audio generation makes our parameters readily understandable and easily modifiable. \\

\end{enumerate}
Automatic programming of virtual synthesizers has also been a topic of interest. In early 2000s, Interactive Genetic Algorithms (IGA's) were utilized for the generation of new sounds with various sound-engines \cite{johnson1999exploring,dahlstedt2001creating}. More recent work by Yee-King et al. \cite{yee2018automatic} used LSTM and genetic algorithms to find the exact parameters used to create a group of sounds. The sounds approximated were made by the same virtual synthesizer, not an external source; making the eventual replication certain, even with random search. Since this work more focused on pads and textures rather than drums, feature matching in this also appears to not be concerned with the envelope of the sounds, but rather the frequency content within arbitrary time windows. Yet another recent, impressive work by Esling et al. used a large dataset of over 10,000 presets for a commercial VST synthesizer to learn a latent parameter space which can be sampled for creation of new audio \cite{esling2019universal}. As stated before, our work presents a rapid approximation of percussion sounds with no previous knowledge about the sonic capabilities of our virtual synthesizer, exploring the actual parameter space rather than its latent representation. 
\section{Implementation}
\label{impl}
% \begin{figure}[H]
% \centering
% \includegraphics[width=0.9\textwidth]{images/SSS_gen.png}
% \caption{Generation Pipeline}
% \label{fig:SSS generator}
% \end{figure}

\subsection{Virtual Synthesizer}
We used the python based Pippi library for sound generation\footnote{https://github.com/luvsound/pippi}. This library uses a C back-end and focuses on fast offline generation of audio signals. In our implementation, a synthesizer can have any number of submodules. We call this number the \textit{stack size}. See table \ref{table:1} for the range of parameters each submodule can take. We call the sets of parameters that characterize a Synth's submodules a \textit{program} (analogous to a preset).  Each submodule can make an audio signal with the length of 0.1-1 second. If a Synth has a stack size of more than 1, these audio signals are overlapped and the total amplitude is normalized.\\  
\begin{table}
\begin{tabular}{ |c|c|c| } 
\hline
Parameters & Value Range & notes and constraints\\
\hline \hline
Attack & 0-3 & A-D-S-R values relative\\
Decay & 0-3 & relative to A-S-R\\
Sustain & 0-3 & relative to A-D-R\\
Release & 0-3 & relative to A-D-S\\
OSC type & sine,square,saw & tone type\\
IsNoise & boolean & whether to generate noise\\
Sample Length & 0-1 second & - \\
StartTime & 0-1 second & Length+Start<1\\
Amplitude & 0.1-1 & 1 = max amplitude\\
Pitches(notes) & list of pitches &  range of C0(16.35hz) to B9 \\
HP filter Cuttoff & 0-20000 & -\\
LP filter Cuttoff & HP-20000 & never lower than HP cutoff\\
Filter Order & 4,8,16 & butterworth filter order \\
\hline
\end{tabular}
\caption{Synthesizer Submodule Parameters}
\label{table:1}
\end{table}
\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{images/specplot.png}
\caption{Mel-spectrum representation of a randomly generated sample. Bigger stacks give us more complex signals, but we find that ever stacks of size 1 can be powerful with the right parameters. }
\label{fig:spectrum grid}
\end{figure}

\subsection{The Ear}
The most challenging and important component of the project; the ear can be any method of evaluating a piece of audio. The most basic representation of a piece of digital audio is via list of numbers and a sampling-rate. For simplicity, we fix the sampling rate to 41,000 per second across all components of this project. An array representation of audio is the representation of the audio within the time domain. More commonly, the frequency domain is used for MIR tasks.
% \begin{figure}[H]
% \centering
% \includegraphics[width=\textwidth]{images/SSS_ear.png}
% \caption{Evaluation Using the Ear}
% \label{fig:SSS generator}
% \end{figure}

All we require from the ear is to give us probabilities of an audio sample belonging to various sound (drum and none-drum) categories. It should be self-evident that these probabilities should sum up to 1.\\

We have implemented several methods of feature extraction thus far and successfully separated different drum categories in interesting ways, visualized in the "interactive\_separation\_graph.ipynb" notebook. A common pitfall of most feature extraction methods implemented is that they do not consider the temporal trends within the audio, making them highly susceptible to errors when evaluating audio that should not belong to any of the categories.\\
\todo[inline, color=green!40]{
Sanity Check: If a drum sample is reversed or chopped up, its likelihood of being labeled the same should change.}

There are two interconnected temporal trends to consider: 
\begin{itemize}
    \item Change in amplitude within each frequency bin. Harder to measure and utilize and possibly not as important for our purposes, considering the short length of our samples.
    \item Overall amplitude change within a sample. Closely linked to ADSR. 
\end{itemize}{}

Temporal trends are can be deduced from a spectrum representation of sound (at least visually by a human). A convolution net trained on our drum dataset led to an ear good enough to continue to the next stage of the project. The current model does an adequate job of rejecting junk data and semi-robustly rejects samples with inappropriate wave-shapes.
\begin{figure}[H]
\centering
\includegraphics[width=0.8\linewidth]{images/cnn_ear.png}
\caption{CNN ear working well on a test set}
\label{fig:ok ear}
\end{figure}

\subsection{Generator+Ear}
Once we have a generator and an ear, we can quickly generate a large dataset where each data item is a random parameter-set, the ear score of the parameter-set and the corresponding rankings for each category. The rankings are derived from the probability scores. For example, the category with the highest probability will have the rank of 1, second highest would rank 2nd, and the lowest probability will have the rank of n (assuming n categories).\\
Important to highlight is that thus far we have only used 1 Synth to generate our sounds, but our generator has the capability to use larger Synth Stacks, and the pipelines are written with that possibility in mind.\\
\todo[inline]{A seen in Figure \ref{fig:rank portions}, there is a bias in random generation towards making higher ranking hats and kicks and lower ranking shakers, claps. Is this bias because of Synth structure or a problem with the ear?}
\begin{figure}[h!]
\centering
\includegraphics[width=1\linewidth]{images/random_ranks.png}
\caption{Figure showing proportions of categorical ranks for around 50,000 randomly generated sounds. Only showing ranks 1-5. About 90\% received the rank of 1 (blue portion of the bar) for the "synth" category, meaning that the ear correctly gave the highest probability to the sound being from a synth combination. We are interested in the small percentages of random generations that trick the ear due to coincidentally being similar to actual drums (i.e those who received a rank of 1) }
\todo[inline, color=green!30]{When looking for the best sounds of each category, could use ranks 2 and 3 etc. as well, however:
\unexpanded{\unexpanded{
\begin{itemize}
    \item We must figure out how to penalize their parameters
    \item The task requires even more confidence in ear,
    \item Generation is fast enough that we don't need to use ranks higher than 1. We have more than enough data if we don't want to deal with this task.
\end{itemize} 
}}}

\label{fig:rank portions}
\end{figure}
\section{Novel Generations}
\subsection{Goal}
 Given a category, our aim is to use random generations to guide an algorithm towards novel generation of drums for that category.

>>>>

\usepackage{float}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{todonotes}
\usepackage{etoolbox}
\usepackage[inline]{enumitem}
\usepackage{subcaption}

\begin{document}

@@

\section{Introduction}
\subsection{Terminology}
\todo[inline,color=green!30]{I'd like to define terms such as VSTs, audio effects, Synthesizers, paramters and presets here. I throw these terms around a lot and I think some might be confused by their meaning. Is this the appropriate place for that? I'd like the read to know these definionts before reading the intro. I can try to keep it brief and engaging}
\subsection{Background and Motivation}
The rise of Digital Audio Workstations (DAW) \cite{leider2004digital} and Virtual Studio Technology (VST) based plug-ins \cite{tanev2013virtual} have rapidly transformed the sonic and material landscape of music production in the recent years. Coupled with this rise in popularity is a vast array of commercial products and services dedicated to satiating the need of amateur and professional music producers for unique sounds; most commonly via audio samples: one-shot drum samples, long sustained notes (commonly referred to as pads or textures), and loops(percussive or melodic) are common deliverables. Two notable examples of these commercial services are \textit{loopmasters}\footnote{loopmasters.com} and \textit{splice.com}\footnote{splice.com}. Furthermore, VST plug-ins can emulate complex audio synthesizers and effects, which some producers may find daunting or time consuming to work with from scratch. In many cases, VST plug-in vendors or unaffiliated enthusiasts sell additional presets for these plugins, targeted towards producers who do not have the time or interest in creating their own. Producers may further modify these presets until their desired sound is reached.\\
Our work is motivated by the idea of finding new, convenient methods for the expansion of a music producer's library of sounds. Primarily with generation of novel, one-shot audio samples but also with automated search and creation of presets for virtual synthesizers that a producer might find  interesting. Using the generation of short, percussive audio samples as a starting point, this project is a proof of concept for promising avenues towards our motivational goal.\\

@@

\subsection{Related Work and Contradistinctions}
\label{related}
Numerous deep, neural network models have been proposed and utilized for the purpose of signal generation in recent years. WaveGans and WaveNet have been subject to significant improvements and experiments since their proposal \cite{nsynth2017}\colorbox{green!=40}{will add more references} and recently, Variational AutoEncoders (VAE's) have been utilized for generation of short percussive samples \cite{aouameur2019neural,ramires2019timbfeat}. In this work however, we opt to use digital signal processing methods to create a virtual synthesizer for the generation of audio signals, as it provides several unique advantages:
\begin{enumerate}[label=\roman*]
  \item Fast, offline rendering of audio with no relience on GPU, currently not possible with state of the art models such as parallel WaveGan \cite{yamamoto2019parallel} and parallell WaveNet \cite{oord2017parallel}. 
  \item Rendering at high sampling rates with little performance issues: Performance speed being a common issue, the standard sampling rate in most audio generation work utilizing neural networks appears to be under 24 khz \cite{yamamoto2019parallel,oord2017parallel,aouameur2019neural,ramires2019timbfeat}. However, a significant number of untrained human ears can detect a change in quality of audio between sampling rates of 192 khz and the industry standard of 44.1 khz \cite{reiss2016meta}, with a dramatic increase in quality detection after training. Therefore we can safely assume that most producers would prefer their audio samples have sampling rates of 44.1 khz or higher. 
  \item Neural networks are often viewed as unexplainable black box solutions. Some models such as VAE's can learn an underlying latent space of parameters and capture the "essence" of the different labels in a dataset. However, these spaces are learned in an unsupervised manner and must be manually analysed, perhaps extensively, before they can be understood \cite{esling2018generative}. The use of a virtual synthesizer for audio generation makes our parameters readily understandable and easily modifiable. \\

\end{enumerate}
Automatic programming of virtual synthesizers has also been a topic of interest. In early 2000s, Interactive Genetic Algorithms (IGA's) were utilized for the generation of new sounds with various sound-engines \cite{johnson1999exploring,dahlstedt2001creating}. More recent work by Yee-King et al. \cite{yee2018automatic} used LSTM and genetic algorithms to find the exact parameters used to create a group of sounds. The sounds approximated were made by the same virtual synthesizer, not an external source; making the eventual replication certain, even with random search. Since this work more focused on pads and textures rather than drums, feature matching in this also appears to not be concerned with the envelope of the sounds, but rather the frequency content within arbitrary time windows. Yet another recent, impressive work by Esling et al. used a large dataset of over 10,000 presets for a commercial VST synthesizer to learn a latent parameter space which can be sampled for creation of new audio \cite{esling2019universal}. As stated before, our work presents a rapid approximation of percussion sounds with no previous knowledge about the sonic capabilities of our virtual synthesizer, exploring the actual parameter space rather than its latent representation. 
\begin{table}[h!]
\centering
\resizebox{\columnwidth}{!}{\begin{tabular}{ |c|c|c| } 
\hline
Parameters & Value Range & notes and constraints\\
\hline \hline
Attack & 0-3 & A-D-S-R values relative\\
Decay & 0-3 & relative to A-S-R\\
Sustain & 0-3 & relative to A-D-R\\
Release & 0-3 & relative to A-D-S\\
OSC type & sine,square,saw & tone type\\
IsNoise & boolean & whether to generate noise\\
Sample Length & 0-1 second & - \\
StartTime & 0-1 second & Length+Start<1\\
Amplitude & 0.1-1 & 1 = max amplitude\\
Pitches(notes) & list of pitches &  range of C0(16.35hz) to B9 \\
HP filter Cuttoff & 0-20000 & -\\
LP filter Cuttoff & HP-20000 & never lower than HP cutoff\\
Filter Order & 4,8,16 & butterworth filter order \\
\hline
\end{tabular}}
\caption{Synthesizer Submodule Parameters}
\label{table:1}
\end{table}

\section{Implementation}
\label{impl}

\subsection{Virtual Synthesizer}
We used the python based Pippi library for sound generation\footnote{https://github.com/luvsound/pippi}. This library uses a C back-end and focuses on fast offline generation of audio signals. In our implementation, a synthesizer can have any number of submodules. We call this number the \textit{stack size}. See table \ref{table:1} for the range of parameters each submodule can take. We call the sets of parameters that characterize a Synth's submodules a \textit{program} (analogous to a preset).  Each submodule can make an audio signal with the length of 0.1-1 second. If a Synth has a stack size of more than 1, these audio signals are overlapped and the total amplitude is normalized.\\

\begin{figure}[h!]
\centering
\begin{subfigure}[b]{\linewidth}
\includegraphics[width=0.3\linewidth]{example-image-a}
\includegraphics[width=0.3\linewidth]{example-image-b}
\includegraphics[width=0.3\linewidth]{example-image-c}
\caption{64 stack}\label{fig:64stack}
\setcounter{subfigure}{0}%
\end{subfigure}

\begin{subfigure}[b]{\linewidth}
\includegraphics[width=0.3\linewidth]{example-image-a}
\includegraphics[width=0.3\linewidth]{example-image-b}
\includegraphics[width=0.3\linewidth]{example-image-c}
\caption{8 stacks}\label{fig:8stacl}
\setcounter{subfigure}{1}%
\end{subfigure}

\begin{subfigure}[b]{\linewidth}
\includegraphics[width=0.3\linewidth]{images/specplot.png}
\includegraphics[width=0.3\linewidth]{images/specplot.png}
\includegraphics[width=0.3\linewidth]{images/specplot.png}
\caption{1 stack}\label{fig:1stack}
\setcounter{subfigure}{2}%
\end{subfigure}
\caption{Examples of mel-spectrum representations of randomly generated samples. Bigger stacks give us more complex signals, but we find that even stacks of size 1 can be powerful with the right parameters. \colorbox{green!=40}{update this with actual spectrums}}
\label{fig:stackspectrums}
\end{figure}

\subsection{The Ear}
We defined an ear as any method of evaluating a piece of audio, capable of "listening" to a piece of audio and giving it a score (or a list of scores) based on how well it satisfies certain criteria. Since in this work we are mainly focused on percussive generation, what we require from the ear is to give us probabilities of an audio sample belonging to various sound (drum and none-drum) categories. As our synthesizer outputs are deterministic for all programs, allowing us to associate the categorical probabilities of each sound with the program that generated it; navigating our synthesis towards parameters which give us the desired sonic output.\\
% \begin{figure}[H]
% \centering
% \includegraphics[width=\textwidth]{images/SSS_ear.png}
% \caption{Evaluation Using the Ear}
% \label{fig:SSS generator}
% \end{figure}
%  The most basic representation of a piece of digital audio is via list of numbers and a sampling-rate. For simplicity, we fix the sampling rate to 41,000 per second across all components of this project. An array representation of audio is the representation of the audio within the time domain. More commonly, the frequency domain is used for MIR tasks.
We have implemented several methods of feature extraction and successfully separated different drum categories in interesting ways. 
\todo[inline, color=green!30]{is this the right place to show off some t-sne graphs of how various feature extraction methods seperated drum categories?}
A common pitfall of most feature extraction methods implemented is that they do not consider the temporal trends within the audio, making them highly susceptible to errors when evaluating audio that should not belong to any of the categories.\\
Analysis of the frequency content of a piece of audio using windowing functions is a common MIR procedure. Care must be taking such that information critical to percussion categorization such as envelope and temporal changes are not lost in this process. For example, we do not want a reversed kick sample to be labeled as a kick, or even a drum, even if it has the correct frequency content. There are two interconnected temporal trends to consider: 
\begin{itemize}
    \item Change in amplitude within each frequency bin. Harder to measure and utilize and possibly not as important for our purposes, considering the short length of our samples.
    \item Overall amplitude change within a sample. Closely linked to ADSR\\. 
\end{itemize}{}
Temporal trends may be deduced from a spectrum representation of sound. A convolution net trained on our drum dataset, combined with an algorithm ADSR extraction algorithm led to an ear which can reliably reject junk data and samples with inappropriate wave-shapes. \todo[inline,color=green!30]{This paragraph is lying, I haven't done the ADSR algorithm part yet. Highest priority after writing this paper.}
\begin{figure}[H]
\centering
\includegraphics[width=0.8\linewidth]{images/cnn_ear.png}
\caption{Model working well on a test set}
\label{fig:ok ear}
\end{figure}
\todo[inline,color=green!30]{ will add f-measure calculations and other confidence inducing graphs (for example I have a grid of spectrums with the ear's guess vs the actual label under each spectrum, I can colorize the grid depending on whether the guess was correct or not). Code in the notebooks. After updating ear, I will make them look presentable and add them.}
\subsection{How Can The Synth and Ear Interact?}
With our Synth and Ear in place, we can quickly generate programs for our Synth, create the corresponding audio sample and get the scores for the audio sample. If the ear returns a vector of categorical probabilities, we can assign categorical ranks to the audio sample. For example, the category with the highest probability will have the rank of 1, second highest would rank 2nd, and the lowest probability will have the rank of n (assuming n categories).\textbf{
But how can we use this to create percussion sounds?}\\\\
We utilize our tools in three different methods of generation, with their specific imlpementation discussed in section \ref{gens}.
\begin {enumerate} [label=(\roman*)]
\item Random Generation: We randomly create a drum programs until a sound belonging to our category is produced. We hypothesize that due to the speed of our experiments, it is a viable approach.
\item Multivariate Distribution Sampling: We attempt to learn the joint distribution of our parameters in order to maximize generation of a certain drum category. 
\item Genetic Search: We treat our parameters as genes and attempt to use evolutionary search algorithms to find programs that the ear deems most fit.\\
\end {enumerate} 
\todo[inline,color=blue!30]{As seen in Figure \ref{fig:rank portions}, there is a bias in random generation towards making higher ranking hats and kicks and lower ranking shakers, claps. Is this bias because of Synth structure or a problem with the ear? Will analyse further after updating ear}
\begin{figure}[h!]
\centering
\includegraphics[width=1\linewidth]{images/random_ranks.png}
\caption{Figure showing proportions of categorical ranks for around 50,000 randomly generated sounds. Only showing ranks 1-5. About 90\% received the rank of 1 (blue portion of the bar) for the "synth" category, meaning that the ear correctly gave the highest probability to the sound being from a synth combination. We are interested in the small percentages of random generations that trick the ear due to coincidentally being similar to actual drums (i.e those who received a rank of 1) }
\todo[inline, color=blue!30]{When looking for the best sounds of each category, could use ranks 2 and 3 etc. as well, however:
\unexpanded{\unexpanded{
\begin{itemize}
    \item We must figure out how to penalize their parameters
    \item The task requires even more confidence in ear,
\end{itemize} 
}}}

\label{fig:rank portions}
\end{figure}
\section{Novel Generations}
\label{gens}
\subsection{Goal}
 Given a category, our aim is to use random generations to guide an algorithm towards novel generation of drums for that category.
